Dark Control:\\ A Unified Account of Default Mode Function\\ by Control Theory and Reinforcement Learning

### 2017-04-19 paper/paper.tex
stuffy stuff

The present paper proposes a \textit{process model}
that tries to explain \textit{how}
the DMN may implement
continuous assessment and prediction of the environment to guide action choices.
%
DMN function is recast in mathematical terms of
control theory and reinforcement learning based on Markov decision processes.
%
We argue that our formal account of DMN function naturally accommodates as special cases
previously proposed cognitive accounts on
\textit{i}) predictive coding,
\textit{ii}) semantic association, and
\textit{iii}) a ``radar'' function.
%
A process model for the neural optimization of complex behavior
can offer parsimonious explanations for
various experimental findings in animals and humans.
\end{abstract}

@@

While sensory experience is a generative process from both views,
the predictive coding account frames
sensory perception of the external world
as a generative experience due to the modulatory top-down influence at
various stages of sensory input processing.
In our RL view, instead,
the Markov property of our sequential decision making framework entails
the environmental model to be
\textbf{effectively expressed in the current state of the human agent
XXX guys, please double-check}.
Further,
the hierarchical processing aspect from predictive coding
is re-expressed in our account in form of
nested prediction of probable upcoming actions, states, and rewards.
Both accounts model the consequences of action.
The predictive coding account is typically explained without
explicit parameterization of the agent's horizon and
has a tendency to be presented as emphasizing prediction about the
immediate future.
In the present account, the horizon of that
look into the future is explicitly expressed in the $\gamma$ parameter
of the Bellman equation.
% \item \suggestremove{Mismatch negativity of PC is immanent in reward prediction error
% by the difference between actually predicted reward of an action given
% a state and the reward predicted by non-linear regression with graded
% discounting of future rewards.}
Finally,
the process of adapting the neuronal connections
for improved top-down modulation
takes the concrete form of gradient computation and
back-propagation in our RL implementation.
It is however important to note that
the neurobiological plausibility of
the back-propagation procedure is controversal
\citep{goodfellow2016deep}.
In sum,
the present formal account may serve as a concrete implementation of
the long-standing predictive coding account from cognitive neuroscience.
MPDs have the advantage of exposing an explicit
mechanism for the horizon of future considerations or
how the internal model of the world gets updated,
and
more explicitly reconciles sensory input processing and action consideration.
%
Recasting DMN function in terms of MDPs therefore naturally incorporates
a majority of aspects of the prediction coding hypothesis.

\subsection{Semantic hypothesis}
Another frequently proposed cognitive account of DMN function revolves
around forming logical associations and analogies between
the current experience and
the conceptual knowledge derived from past experiences
\citep{bar2007proactive, binder1999conceptual}.
Analogies might naturally tie incoming novel sensory stimuli to
explicit world knowledge (i.e., semantics) extracted from the environment
\citep{bar2009proactive}.
The encoding of complex environments could thus be facilitated
by association to similar states.
%
Going beyond human language itself,
semantic building blocks may provide the basis for
mentally envisioning non-existent scenarios
that would optimize behavior in the environment
by constantly simulating possible future events.
Such cognitive processes can afford
the internal generation of necessary information
that is not presented in the surrounding environment
by recombining building blocks of
concept knowledge and episodic memories
\citep{hassabis2009construction}.
Indeed, in aging humans, remembering the past and imaging the future
equally decreased in the level of detail and were associated with
concurrent deficits in forming and integrating relationships between
items \citep{addis2008age, spreng2006temporal}.
Further,
a constructive account can explain the reciprocal relationship
between an egocentric first person perspective and
an allocentric bird’s eye perspective immersed in
self-reflection, semantic associations, and autobiographical memories.
%
Cognitive aspects of egocentric-allocentric switching
are also closely related to episodic memory, language, problem solving,
planning, estimating other people's thoughts, and spatial navigation.
These neural processes are likely to
build on abstract world knowledge and logical assocations
for integrating the constituent elements in mental scene construction
\citep{schacter2007remembering}.
Such scene generation processes could contribute to interpreting the
present and foretelling the future.
Mental scene construction has been proposed
to imply a distinction between
engagement in the sensory environment
and internally generated mind-wandering
\citep{buckner2007self}.
These investigators stated that
``A computational model [...] will probably require a form of
regulation by which perception of the current world is suppressed
while simulation of possible alternatives are constructed,
followed by a return to perception of the present.''.
The semantic hypothesis is for instance supported by evidence in animals that
could learn a \textit{cognitive map of the environment},
even without reward incentives, and exploit it later
for other means \citep{tolman1948cognitive}.

In comparison,
both the semantic hypothesis and MDPs expose mechanisms of
how action considerations could be mentally explored.
In both accounts,
there is also no evidence to indicate that predictions of various
levels of complexity, abstraction, timescale and purpose
use mechanisms that are qualitatively different. This concurs with
DMN activity increases across time, space, and content domains
demonstrated in many neuroimaging studies \citep{spreng2009common}.
The semantic hypothesis
and MDP account provide explanations why hippocampus lesion does
not only impair recalling memories, but also hypothetical and future
thinking \citep{hassabis2007patients}.
While both semantic hypothesis and
our formal account propose memory-based internally
generated information for probabilistic mental models of action outcomes,
MDPs render explicit the grounds on which the final action is
eventually chosen, namely, the estimated cumulative reward.
\textbf{The notion of semantic knowledge associations is
incorporated into the MDP as the Markov property,
that is, the current state directly results from the
agent's history of states and actions. The learned
estimations of value and action transitions drive
stimulus processing and action choice in the present.}
The MDPs naturally integrate the former egocentric
(more related to current action, state and reward) and the later
allocentric (more related to past and future actions, states, and rewards)
angles on the world in a same optimization problem.
Further,
the semantic account of DMN function does not offer
a mechanistic explanation how explicit world knowledge and semantic analogies thereof
lead to prediction of future actions and states.
The semantic hypothesis does not explain why memory recall
for scene construction in humans is typically fragmentary and noisy
instead of accurate and reliable.
In contrast to existing accounts on semantics and
mental scene construction, the random and creative aspects of DMN function
are explained in MDPs by the advantages of stochastic optimization.
Yet, the MDP-based account provides an algorithmic explanation in that
stochasticity of the parameter space explored
by the Monte Carlo solvers achieves better fine-tuning of the
action policies and estimation of expected reward outcomes.
That is, the purposeful stochasticity of policy and value estimation
in MDPs provides a candidate explanation for why humans
have evolved imperfect noisy memories
as the more advantageous adaptation.
In sum, episodic scene construction according to the semantic
account is lacking an explicit time and incentive model.

\subsection{Sentinel hypothesis}
Processing self-relevant information was perhaps the first
cognitive account that was proposed for the DMN \citep{gusnard2001medial}.
Since then,
many investigators have speculated that neural activity in the DMN
may reflect the brain's continuous tracking of
relevance in the environment
as an advantageous evolutionary adaptation \citep{randy2008, hahn2007cingulate}.
According to this cognitive account, the human brain's baseline realizes
a ``radar'' function to
detect subjectively salient cues and unexpected events in the environment.
Propositions of such a sentinel function have however seldom detailed
the mechanisms of
how attention and memory resources are exactly reallocated when
encountering a salient environmental stimulus.
In the present formal account,
promising action trajectories
are recursively explored by human agents. Conversely,
certain branches of candidate action trajectories
are detected to be less worthy to become mentally explored.
This mechanism, expressed by the Bellman equation,
directly implies stratified allocation of attention and working memory load
over relevant cues and events in the environment.
%
Further,
our account provides a parsimonious explanation for
the increasingly recognized observation
of the DMN implication
in certain goal-directed experimental tasks and
in task-unconstrained mind-wandering \citep{smith2009, bzdok2016formal}.
Both cognitive sets may entail DMN recruitment
if environmental relevance is processed, manipulated, and
used for predicting the future.
During tasks,
the policy and value estimates may be
updated to optimize especially short-term action.
At rest, these parameter updates may
improve especially mid- and long-term action.
The horizon is expressed in the $\gamma$ parameter in the formal account.
We thus provide answers for the unsettled question why the involvement
of the same neurobiological brain circuit (i.e., DMN) has been documented
for baseline house-keeping functions and specific task performances.

@@

\section{Conclusion}
What brain function could be important enough
for the existence and survival of the human species
to warrant constantly high energy costs?
MDPs provide an attractive
formal account how the human association cortex
might implement supra-modal representation and control of the environment to
optimize the organism's fate.
% by obtaining reward and avoiding pain.
This idealized process model explains
a number of previous experimental observations in the
DMN by a simple but non-trivial mechanism.
%
From the computational view of a Markovian sequential decision process,
behavior unfolds by integrating happened past events
and possible future events to guide action choice in the present context.
MDPs provide a mathematical formalism how
optimal substructure in the environment can be recursively exploited
when confronted with complicated decisions.
%
This functional interpretation is more compatible with the DMN's
poorly understood involvement across
autobiographical memory recall, problem solving,
abstract reasoning based on internally generated scenes, social cognition,
as well as delay discounting and self-related prospection.
Improvement of the internal world model
by injecting stochasticity in the recall of past
actions and outcomes may explain why
very accurate memories have been disfavored in human evolution
and why human creativity might be adaptive.
%
In principle,
neuroscience experiments can be designed that operationalize
the set of action, value, and state variables that determine
the behavior of intelligent RL systems.

The proposed machine-learning perspective
on DMN biology is hence not only practically computable but
also yields falsifiable neuroscientific hypotheses.
%
At the least, we propose an alternative vocabulary to
describe and interpret experimental findings in neuroscience studies
on this macroscopic brain network.
%
Ultimately,
the DMN can be viewed as a functional integrator
from the relatively recent events to anticipated upcoming events
in to constantly refine action in our dynamic world.

>>>>

The present paper proposes a \textit{process model}
that tries to explain \textit{how}
the DMN may implement
continuous assessment and prediction of the environment to guide action choices.
%
DMN function is recast in mathematical terms of
control theory and reinforcement learning based on Markov decision processes.
%
We argue that our formal account of DMN function naturally accommodates as special cases
previously proposed cognitive accounts on
\textit{i}) predictive coding,
\textit{ii}) semantic association, and
\textit{iii}) a ``sentinel'' function.
%
A process model for the neural optimization of complex behavior
can offer parsimonious explanations for
various experimental findings in animals and humans.
\end{abstract}

@@

While sensory experience is a generative process from both views,
the predictive coding account frames
sensory perception of the external world
as a generative experience due to the modulatory top-down influence at
various stages of sensory input processing.
In our MDP view, instead,
the Markov property of our sequential decision making framework entails
the environmental model to be
\textbf{effectively expressed in the current state of the human agent
XXX guys, please double-check}.
Further,
the hierarchical processing aspect from predictive coding
is re-expressed in our account in form of
nested prediction of probable upcoming actions, states, and rewards.
Both accounts capture the consequences of action.
The predictive coding account is typically explained without
explicit parameterization of the agent's horizon and
has a tendency to be presented as emphasizing prediction about the
immediate future.
In the present account, the horizon of that
look into the future is explicitly expressed in the $\gamma$ parameter
of the Bellman equation.
% \item \suggestremove{Mismatch negativity of PC is immanent in reward prediction error
% by the difference between actually predicted reward of an action given
% a state and the reward predicted by non-linear regression with graded
% discounting of future rewards.}
Finally,
the process of adapting the neuronal connections
for improved top-down modulation
takes the concrete form of gradient computation and
back-propagation in our MDP implementation.
It is however important to note that
the neurobiological plausibility of
the back-propagation procedure is controversal
\citep{goodfellow2016deep}.
In sum,
the present formal account may serve as a concrete implementation of
the long-standing predictive coding account from cognitive neuroscience.
MDPs have the advantage of exposing an explicit
mechanism for the horizon of future considerations or
how the internal model of the world gets updated,
and
more explicitly reconciles sensory input processing and action consideration.
%
Recasting DMN function in terms of MDPs therefore naturally incorporates
a majority of aspects from the prediction coding hypothesis.

\subsection{Semantic hypothesis}
Another frequently proposed cognitive account of DMN function revolves
around forming logical associations and abstract analogies between
the current experience and
the conceptual knowledge derived from past experiences
\citep{bar2007proactive, binder1999conceptual, constantinescu2016organizing}.
Analogies might naturally tie incoming novel sensory stimuli to
explicit world knowledge (i.e., semantics) extracted from the environment
\citep{bar2009proactive}.
The encoding of complex environments could thus be facilitated
by association to similar states.
%
Going beyond human language itself,
semantic building blocks may provide the basis for
mentally envisioning non-existing scenarios
that would optimize behavior in the environment
by constantly simulating possible future events.
Such cognitive processes can afford
the internal generation of necessary information
that is not presented in the surrounding environment
by recombining building blocks of
concept knowledge and episodic memories
\citep{hassabis2009construction}.
Indeed, in aging humans, remembering the past and imaging the future
equally decreased in the level of detail and were associated with
concurrent deficits in forming and integrating relationships between
items \citep{addis2008age, spreng2006temporal}.
Further,
% a constructive account can explain the reciprocal relationship
% between an egocentric first person perspective and
% an allocentric bird’s eye perspective immersed in
% self-reflection, semantic associations, and autobiographical memories.
% %
% Cognitive aspects of egocentric-allocentric switching
episodic memory, language, problem solving,
planning, estimating other people's thoughts, and spatial navigation
represent neural processes that are likely to
build on abstract world knowledge and logical assocations
for integrating the constituent elements in mental scene construction
\citep{schacter2007remembering}.
Such scene generation processes could contribute to interpreting the
present and foretelling the future.
Mental scene construction has been proposed
to imply a distinction between
engagement in the sensory environment
and internally generated mind-wandering
\citep{buckner2007self}.
These investigators stated that
``A computational model [...] will probably require a form of
regulation by which perception of the current world is suppressed
while simulation of possible alternatives are constructed,
followed by a return to perception of the present.''.
The semantic hypothesis is for instance supported by evidence in animals that
could learn a \textit{cognitive map of the environment}
and exploit it later for other means \citep{tolman1948cognitive}.

In comparison,
both the semantic hypothesis and the present account based on MDPs
expose mechanisms of how action considerations could be mentally explored.
In both accounts,
there is also no reason to assume that predictions of various
levels of complexity, abstraction, timescale and purpose
use mechanisms that are qualitatively different. This concurs with
DMN activity increases across time, space, and content domains
demonstrated in many neuroimaging studies
\citep{spreng2009common, laird2009, bzdok2012morality, binder2009}.
The semantic hypothesis
and MDP account provide explanations why hippocampus lesion does
not only impair recalling memories, but also hypothetical and future
thinking \citep{hassabis2007patients}.
While both semantic hypothesis and
our formal account propose memory-based internally
generated information for probabilistic mental models of action outcomes,
MDPs render explicit the grounds on which the final action is
eventually chosen, namely, the estimated cumulative reward.
\textbf{The notion of semantic knowledge associations is
incorporated into the MDP as the Markov property,
that is, the current state directly results from the
agent's history of states and actions. The learned
estimations of value and action transitions drive
stimulus processing and action choice in the present.}
The MDPs naturally integrate the former egocentric view
(more related to current action, state and reward) and the later
world view (more related to past and future actions, states, and rewards)
on the world in a same optimization problem.
Further,
the semantic account of DMN function does not offer
a mechanistic explanation how explicit world knowledge and semantic analogies thereof
lead to prediction of future actions and states.
The semantic hypothesis does not explain why memory recall
for scene construction in humans is typically fragmentary and noisy
instead of accurate and reliable.
In contrast to existing accounts on semantics and
mental scene construction, the random and creative aspects of DMN function
are explained in MDPs by the advantages of stochastic optimization.
Yet, the MDP-based account provides an algorithmic explanation in that
stochasticity of the parameter space explored
by the Monte Carlo solvers achieves better fine-tuning of the
action policies and estimation of expected reward outcomes.
That is, the purposeful stochasticity of policy and value estimation
in MDPs provides a candidate explanation for why humans
have evolved imperfect noisy memories
as the more advantageous adaptation.
In sum, episodic scene construction according to the semantic
account is lacking an explicit time and incentive model.

\subsection{Sentinel hypothesis}
Processing self-relevant information was perhaps the first
cognitive account that was proposed for the DMN \citep{gusnard2001medial}.
Since then,
many investigators have speculated that neural activity in the DMN
may reflect the brain's continuous tracking of
relevance in the environment
as an advantageous evolutionary adaptation \citep{randy2008, hahn2007cingulate}.
According to this cognitive account, the human brain's baseline realizes
a ``radar'' function to
detect subjectively salient cues and unexpected events in the environment.
Propositions of such a sentinel function have however seldom detailed
the mechanisms of
how attention and memory resources are exactly reallocated when
encountering a salient environmental stimulus.
In the present formal account,
promising action trajectories
are recursively explored by human agents. Conversely,
certain branches of candidate action trajectories
are detected to be less worthy to become mentally explored.
This mechanism, expressed by the Bellman equation,
directly implies stratified allocation of attention and working memory load
over relevant cues and events in the environment.
%
Further,
our account provides a parsimonious explanation for
the increasingly recognized observation
of the DMN implication
in certain goal-directed experimental tasks and
in task-unconstrained mind-wandering \citep{smith2009, bzdok2016formal}.
Both environment-detached and environment-engaged
cognitive sets may entail DMN recruitment
if environmental relevance is processed, manipulated, and
used for predicting the future.
During tasks,
the policy and value estimates may be
updated to optimize especially short-term action.
At rest, these parameter updates may
improve especially mid- and long-term action.
The horizon is expressed in the $\gamma$ parameter in the formal account.
We thus provide answers for the currently unsettled question why the involvement
of the same neurobiological brain circuit (i.e., DMN) has been documented
for baseline house-keeping functions and specific task performances.

@@

\section{Conclusion}
What brain function could be important enough
for the existence and survival of the human species
to warrant constantly high energy costs?
MDPs provide an attractive
formal account how the human association cortex
might implement supra-modal representation and control of the environment to
optimize the organism's fate.
% by obtaining reward and avoiding pain.
This idealized process model explains
a number of previous experimental observations in the
DMN by simple but non-trivial mechanisms.
%
From the computational view of a Markovian sequential decision process,
behavior unfolds by integrating happened past events
and possible future events to guide action choice in the present context.
MDPs provide a mathematical formalism how
optimal substructure in the environment can be recursively exploited
when confronted with complicated decisions.
%
This functional interpretation is more compatible with the DMN's
poorly understood involvement across
autobiographical memory recall, problem solving,
abstract reasoning, social cognition,
as well as delay discounting and self-related prospection.
Improvement of the internal world model
by injecting stochasticity in the recall of past
actions and the estimation of action outcomes may explain why
very accurate memories have been disfavored in human evolution
and why human creativity might be adaptive.
%
In principle,
neuroscience experiments can be designed that operationalize
the set of action, value, and state variables that determine
the behavior of intelligent RL systems.

The proposed machine-learning perspective
on DMN biology is hence not only practically computable but
also yields falsifiable neuroscientific hypotheses.
%
At the least, we propose an alternative vocabulary to
describe and interpret experimental findings in neuroscience studies
on this major brain network.
%
Ultimately,
the DMN can be viewed as a functional integrator
from the relatively recent events to anticipated upcoming events
in to constantly refine the organism's interaction with the world.
